---
title: "biodogma"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{biodogma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(biodogma)
```

### Description of the package `biodogma`:

-   **Function 1:** `gen_dna_seq`

-   **Function 2:** `dna_to_rna`

-   **Function 3:** `split_codon`

-   **Function 4:** `gen_aa_seq`

-   **Function 5:** `aa_count`

Amino acid sequences contain valuable information about protein structure. The `aa_count` function helps visualize the composition of these sequences, being useful to analyze certain protein characteristicas like the enrichment in certain amino acid.

#### Installation

Make sure to have the `stringr` and `ggplot2` packages installed.

#### Usage

`aa_count(seq)`


##### Arguments

-   `seq`: a character **string** containing the amino acid sequence.

##### Return

-   A `ggplot` bar plot showing the absolute frequencies for each amino acid present in the given sequence.

#### Example

```{r}
library("biodogma")

# Provide an amino acid sequence
seq1 <- "AKLPYNILSDFHVNSTRLGE"

# Call the function 
aa_plot <- aa_count(seq1)
```

### Discussion

-   Use cases for the package and what other functions could be included.

-   Task 3 - Group discussion
