---
title: "biodogma"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{biodogma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(biodogma)
```

## Description of the package `biodogma`:

The package `biodogma` consists of 5 simple functions that allows us to model the central dogma of molecular biology. This theory was proposed by Francis Crick in 1957 and states that genetic information flows from **DNA** to **RNA** via the process of [transcription]{.underline}, and from RNA to **protein** via [translation]{.underline}. To replicate this, the package allows us to generate random DNA sequences, transform them into RNA sequences, split the RNA sequences into codons, translate the codons into amino acids, and finally visualise the amino acid whencontent in a protein sequence.

The functions included in the `biodogma` package are:

### **Function 1:** `gen_dna_seq()`

The `gen_dna_seq` function generates a random DNA sequence of a specified length. Each nucleotide in the sequence is chosen randomly from the set {A, T, G, C} to simulate a DNA sequence.

#### **Usage**

`gen_dna_seq(length)`

##### **Arguments**

-   `length`: An integer specifying the desired length of the DNA sequence. This value determines the number of nucleotides in the generated sequence.

##### **Return**

-   A character string representing a randomly generated DNA sequence of the specified length, consisting of the nucleotides "A" (adenine), "T" (thymine), "G" (guanine), and "C" (cytosine).

#### Example

```{r}
# Load the biodogma package
#library(biodogma)

# Define the desired length for the DNA sequence
length <- 10

# Generate a random DNA sequence of the specified length
dna_seq <- gen_dna_seq(length)

# Print the generated DNA sequence
print(dna_seq)

```

#### Step-by-Step Explanation

-   The function uses the `sample()` function to randomly select characters from the set `c("A", "T", "G", "C")`. The `size` argument, set to `length`, specifies the number of nucleotides to generate, and `replace = TRUE` allows repeated selection of any nucleotide.

-   The randomly selected nucleotides are combined into a single string using `paste0()` with `collapse = ""`, creating a contiguous DNA sequence without spaces between characters.

### **Function 2:** `dna_to_rna()`

In the process of transcription, DNA is converted to RNA by replacing each occurrence of thymine (T) with uracil (U), which is specific to RNA. The purpose of `dna_to_rna` is to substitute "T" with "U" in a DNA sequence.

#### Usage

`dna_to_rna(dna_sequence)`

##### Arguments

-   `dna_sequence`: a character string containing the DNA sequence that needs to be transcribed. For example, a valid input might be "ATGCTTAC", where each letter represents a nucleotide base (adenine, thymine, guanine, cytosine).

##### Return

-   A character string representing the RNA sequence, with all occurrences of "T" replaced by "U". For example, the input "ATGCTTAC" would return "AUGCUUAC".

#### Example

```{r}
# Load the biodogma package
#library("biodogma")  

# Define a DNA sequence
dna_sequence <- "ATGCTTAC"

# Transcribe DNA to RNA
rna_sequence <- dna_to_rna(dna_sequence)

# Print the RNA sequence
print(rna_sequence)
```

#### Step-by-step explanation

The `dna_to_rna` function uses the `gsub()` function to perform a global substitution on the input DNA sequence. Specifically, "T" characters are replaced with "U", simulating the transcription process where thymine in DNA is substituted with uracil in RNA.

### **Function 3:** `split_codon()`

mRNA sequences are read in codons or groups of three ribonucleotides, each of these coding for a specific amino acid. The purpose of `split_codon` is to split a string of an RNA sequence into a vector of codons.

#### Usage

`split_codon(rna_seq, starting_codon_position)`

##### Arguments

-   `rna_seq`: a character string containing the RNA sequence.

-   `starting_codon_position`: an int variable providing the position of the initial codon in the given sequence. It's *not necessary* to provide this argument, the default value being position 1.

##### Return

-   A vector containing the codons found in the given string.

#### Example

```{r}
# Load the biodogma package
#library("biodogma")

# Provide an RNA sequence
rna_seq <- "AUGACUAAGA"
pos <- 2

# Call the function without specifying the starting position
split_codon(rna_seq)

# Call the function specifying the starting position
split_codon(rna_seq, pos)
```

#### Step-by-step explanation

The function `split_codon` follows these steps:

1.  Obtains the length of `rna_seq` using `nchar()` and saves it in a variable (`len`).

Splits `rna_seq` into codons or non-overlapping 3-character strings, using the function `susbtring()`. In the function `susbtring()` the arguments given are the `rna_seq` string and sequences of the initial and ending positions of each codon. These sequences are defined using `seq()`, which calls both `len` and `starting_codon_position`.

### Function 4: `gen_aa_seq()`

After the mRNA has been synthesized, the molecule leaves the nucleus and is read by ribosomes for generating the resulting protein. In this process, called biological translation, each triplet of ribonucleotides ('codons') result in one specific amino acid, that will generate the amino acid chain. This equivalence happens accordingly to the genetic code.

**Usage**

`gen_aa_seq <- function(codons)`

##### **Arguments**

codon: a vector containing the triplets of the considered mRNA chain.

##### **Return**

After calling the function, the user will get an amino acid chain or polypeptide.

**Example**

```{r}
#Load the library
#library("biodogma")

#Define a vector with the codons that will be translated into proteins. 
#NOTE: the user may obtain this vector from the function splitcodon(). 
codons <-  c("AUG", "ACU", "AAG") 

#Call the function 
aa_seq <- gen_aa_seq(codons)
print(aa_seq)
```

#### Step-by-step explanation

`gen_aa_seq()` receives a vector with the codons or nucleotide triplets from a given mRNA sequence. Each of them is converted into an amino acid using `genetic_code`. These characters will be concatenated with the `paste0()` function, without any separation.

### **Function 5:** `aa_count()`

Amino acid sequences contain valuable information about protein structure. The `aa_count` function helps visualize the composition of these sequences, being useful to analyze certain protein characteristics like the enrichment in certain amino acid. Thus, the output of the function is a character variable including the amino acid chain.

#### Installation

Make sure to have the `stringr` and `ggplot2` packages installed. If not, you can install them using:

`install.packages("stringr")`

`install.packages("ggplot2")`

#### Usage

`aa_count(seq)`

##### Arguments

-   `seq`: a character **string** containing the amino acid sequence.

##### Return

-   A `ggplot` bar plot showing the absolute frequencies for each amino acid present in the given sequence.

#### Example

```{r}
# Load the biodogma package
#library("biodogma")

# Provide an amino acid sequence
seq1 <- "AKLPYNILSDFHVNSTRLGE"

# Call the function 
aa_plot <- aa_count(seq1)
aa_plot
```

This example produces a barplot saved in the `aa_plot` object.

#### Step-by-step explanation

The function follows these steps:

1.  Obtain a vector with unique amino acids in the sequence.
2.  Counts amino acids using `str_count` from the `stringr` package.
3.  Creates data frame with the frequencies of each amino acid.
4.  Plot: using `ggplot2` a bar plot is generated showing the absolute frequencies.

### Discussion

#### Use Cases for `biodogma`

The `biodogma` package is useful in:

-   **Education**: This package helps students and educators simulate the processes of transcription and translation, visualizing how DNA is converted to RNA and then to proteins.

-   **Bioinformatics Research**: Researchers can use the package for quick simulations, such as generating synthetic DNA/RNA sequences, analyzing codon usage, and visualizing amino acid compositions in protein sequences, which can help in understanding protein structure, function, and enrichment of specific amino acids.

#### Potential Additional Functions

To enhance the package’s capabilities, the following functions could be added:

-   **Reverse Transcription**: A function to convert RNA back into DNA.

-   **Codon Optimization**: A function that suggests codon sequences based on species-specific codon usage bias.

-   **Protein Properties**: Functions to calculate protein properties such as molecular weight, hydrophobicity, or charge, which could assist in protein characterization and function prediction.

#### Dependencies

Although it is usually better to avoid dependencies, in this package we have used functions from `stringr` and `ggplot2` packages, due to their improved functionality and great usage. The avoidance of dependencies simplifies the complexity of the maintenance of packages as new changes in other packages may affect the functionality of another. Furthermore, it increases the usage as it does not need additional packages besides the one you are installing. However, this is not always possible as packages like those integrating the tidyverse implement complex function in an efficient manner.

In this package we have used `@importFrom package function` tags in the function description because it reduces the repetition of the `package::function` form, however the latter does not add it to the namespace and reduces the dependencies as it only loads the used functions and not the whole package (although this can also be achieved by using `min_version = TRUE` in the other form).
